---
title: "ST300 Week 5"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
## Effect of correlated predictors

Suppose that $\tilde{X}_1, \tilde{X}_2$ and $\tilde{X}_3$ are iid $N(0,1)$, and $Z$ is independently distributed as $N\left(0, \sigma_z^2\right)$. Define $X_j=\tilde{X}_j+Z, j=1, \ldots, 3$. What is the correlation between pairs $X_j$ and $X_{j^{\prime}}$?

Suppose we construct a regression function $f(X) = X \hat{\beta}$. where $\hat{\beta}=c(1,1,1)$. What is the variance of $f(X)$ (i.e. unconditional variance)?

Write a function `gendata` that generates data in the following fashion: It takes as input $N, \rho, SNR$ (signal-to-noise ratio), and $RNseed$, a random number seed. It should generate a $N \times 3$ matrix $X$ of correlated predictors, with pairwise correlations $\rho$ and variances 1. It should create a regression function $f$ as described above, and a response $Y = f(X)+\epsilon$ such that $\text{Var}(f(X)) / \text{Var}(\epsilon)=SNR$. It should return $X$ and $y$ as named components in a `dataframe`.
```{r}
gendata <- function(N, rho, SNR, RNseed) {
  # Set the seed for reproducibility
  set.seed(???)
  
  # Generate correlated predictors
  # Correlation matrix for X
  Sigma <- ???
  
  # Generate multivariate normal data for X: use mvrnorm from package MASS
  X <- MASS::mvrnorm(???)
  
  # Define regression function f(X) = X1 + X2 + X3
  f_X <- 
  
  # Calculate variance of f(X)
  var_fx <- 
  
  # Determine the variance of epsilon (noise) based on SNR
  var_epsilon <- var_fx / SNR
  
  # Generate epsilon (noise) from normal distribution with mean zero and variance var_epsilon
  epsilon <- 
  
  # Generate response Y = f(X) + epsilon
  Y <- 
  
  # Return X and Y in a dataframe
  return(data.frame(X1 = X[,1], X2 = X[,2], X3 = X[,3], Y = Y))
}

# Example usage:
# gendata(100, 0.5, 2, 123)

```

Write a function `coefcor` that takes as input the same as the above, except $\rho$ is now a vector of correlations. For each value of $\rho$, it computes the OLS coefficients for $X$, and stores them as rows of a matrix (length $(\rho) \times 3$ ). For each value of $\rho$, it uses the same RN seed.
```{r}
coefcor <- function(N, rho_vec, SNR, RNseed) {
  # Initialize a matrix to store the coefficients (length of rho_vec x 3)
  coef_matrix <- matrix(0, nrow = length(rho_vec), ncol = 3)
  
  # Loop over each correlation in rho_vec
  for (i in seq_along(rho_vec)) {
    rho <- rho_vec[i]  # Current correlation value
    
    df <- # data generated using gendata
    
    # Fit linear model Y ~ X1 + X2 + X3 using OLS for Y, X1, X2, and X3 from df
    lm_fit <- 
    
    # Extract coefficients (ignore the intercept, take X1, X2, X3)
    coef_matrix[i, ] <- 
  }
  
  # Return the coefficient matrix
  return(coef_matrix)
}

# Example usage:
# rho_vector <- c(0.1, 0.3, 0.5, 0.7, 0.9)
# coefcor(100, rho_vector, 2, 123)
```

Run your function `coefcor` using $N = 30$ for 20 values of $\rho$ between 0 and 0.95 , and $\mathrm{SNR}=1$. Plot the coefficients of each variable as a function of $\rho$ (on the same plot). Do this 100 times (with a different seed each time) and accumulate the results (You don't have to save the plots - you can watch them run by).
```{r}
# Parameters
N <- 30
rho_vec <- seq(0, 0.95, length.out = 20)
SNR <- 1
num_iterations <- 100

# Store coefficients over iterations
coefficients_X1 <- matrix(0, nrow = length(rho_vec), ncol = num_iterations)
coefficients_X2 <- matrix(0, nrow = length(rho_vec), ncol = num_iterations)
coefficients_X3 <- matrix(0, nrow = length(rho_vec), ncol = num_iterations)

# Run the simulation 100 times with different random seeds
for (seed in 1:num_iterations) {
  coef_matrix <- coefcor(N, rho_vec, SNR, RNseed = seed)
  
  # Store the results
  coefficients_X1[, seed] <- coef_matrix[, 1]
  coefficients_X2[, seed] <- coef_matrix[, 2]
  coefficients_X3[, seed] <- coef_matrix[, 3]
}

# Plot the results
par(mfrow = c(1, 1))  # Reset plotting layout

# Plot coefficients for X1, X2, X3 as a function of rho
plot(rho_vec, coefficients_X1[,1], type = "l", col = "blue", ylim = range(c(coefficients_X1, coefficients_X2, coefficients_X3)),
     xlab = expression(rho), ylab = "Coefficients", main = "OLS Coefficients vs Correlation (100 Iterations)")
for (i in 2:num_iterations) {
  lines(rho_vec, coefficients_X1[, i], col = "blue")
}
for (i in 1:num_iterations) {
  lines(rho_vec, coefficients_X2[, i], col = "red")
}
for (i in 1:num_iterations) {
  lines(rho_vec, coefficients_X3[, i], col = "green")
}

# Add a legend to differentiate the variables
legend("topright", legend = c("X1", "X2", "X3"), col = c("blue", "red", "green"), lty = 1)
```

Summarize the results as follows. Make a single plot of each coefficient profile as a function of $\rho$ (there should be 300 curves in the plot). For each value of $\rho$, compute the correlation matrix of the coefficients across the 100 runs. Average the three correlations, and plot them as a function of $\rho$. Summarize what you have learned.
```{r}
# Initialize a vector to store the averaged correlations across all rho values
avg_correlations <- numeric(length(rho_vec))

# Loop over each rho and compute correlation matrix of the coefficients
for (i in 1:length(rho_vec)) {
  # Collect the coefficients for X1, X2, X3 across the 100 iterations
  coef_data <- cbind(coefficients_X1[i, ], coefficients_X2[i, ], coefficients_X3[i, ])
  
  # Compute the correlation matrix of the coefficients
  cor_matrix <- cor(coef_data)
  
  # Extract the off-diagonal correlations and average them
  avg_correlation <- mean(c(cor_matrix[1, 2], cor_matrix[1, 3], cor_matrix[2, 3]))
  
  # Store the average correlation for the current rho
  avg_correlations[i] <- avg_correlation
}

# Plot the average correlations as a function of rho
plot(rho_vec, avg_correlations, type = "b", col = "purple", pch = 19, 
     xlab = expression(rho), ylab = "Average Correlation of Coefficients",
     main = "Average Correlation of Coefficients vs Rho")
```

## R-squared
**Exercise 34.** Show that $R^2$ is the squared sample Pearson correlation coefficient between $Y$ and $\hat{Y}$.

## F test
The data for this example come from a study by Stamey et al. (1989). They examined the correlation between the level of prostate-specific antigen and a number of clinical measures in men who were about to receive a radical prostatectomy. The variables are log cancer volume `lcavol`, log prostate weight `lveight`, `age`, log of the amount of benign prostatic hyperplasia `1bph`, seminal vesicle invasion `svi`, log of capsular penetration `lcp`, Gleason score `gleason`, and percent of Gleason scores 4 or 5 `pgg45`. Read the data set and save it as a data frame. Then calculate the pairwise correlations. 

```{r}
prostate <- 

```
Fit a linear model to the log of prostate-specific antigen, `lpsa`, after first standardizing the predictors to have unit variance. Which predictors are not significant at the $5\%$ level? What happens if you exclude all non-significant predictors from regression? Use F-test to justify your reasoning.
```{r}
prostate_scale <- 
fit_all <- 

# which variables are not significant

fit_sub <- 

# use function anova for F-test
```

## Prediction Interval
(2016 exam Q2 b) Let $Y=X \boldsymbol{\beta}+\boldsymbol{\epsilon}, \boldsymbol{\epsilon} \sim N\left(\mathbf{0}, \sigma^2 \mathrm{I}_n\right)$ be a linear model, with data $\left(\mathbf{x}_i, y_i\right), i=1, \ldots, n$. The parameter $\boldsymbol{\beta}$ is $p$ dimensional.
Two new data points are observed : $\left(\mathbf{w}, y_w\right)$ and $\left(\mathbf{z}, y_z\right)$. That is,

$$
y_w=\mathbf{w}^{\mathbf{T}} \boldsymbol{\beta}+\epsilon_w, \quad y_z=\mathbf{z}^{\mathbf{T}} \boldsymbol{\beta}+\epsilon_z,
$$

where $\epsilon_w, \epsilon_z \sim N\left(0, \sigma^2\right)$ and are independent of each other, and are independent from all past data also. Suppose $\mathbf{w}, \mathbf{z}$ are both known covariate vectors.
i. Show that the distribution of $y_w-y_z$ is

$$
y_w-y_z \sim N\left((\mathbf{w}-\mathbf{z})^{\mathbf{T}} \boldsymbol{\beta}, 2 \sigma^2\right) .
$$

ii. Derive the distribution of $\left(\hat{y}_w-\hat{y}_z\right)-\left(y_w-y_z\right)$.
iii. Construct a $(1-\gamma)$ prediction interval for $y_w-y_z$.
iv. Suppose now that $\mathbf{w}$ and $\mathbf{z}$ are independent normal with mean $\mathbf{0}$ and covariance matrix $\sigma_x^2 \mathbf{I}$, with $\sigma_x^2$ known. Find the distribution of $y_w-y_z$ in terms of $\boldsymbol{\beta}, \sigma_x^2$ and $\sigma^2$. Suggest an approximate $(1-\gamma)$ prediction interval for $y_w-y_z$ by estimating its variance using a reasonable plug-in estimator.

